---
title: "cp_ml.Rmd"
author: "Sunny Khosla"
date: "Saturday, July 25, 2015"
output: html_document
---
```{r, Global settings, echo=FALSE} 
            library(curl) # on R console do install.packages("curl")
            library(caret) # on R console do install.packages("caret") & install.packages('e1071', dependencies=TRUE)
            library(rpart)
            library(e1071)
```



SUMMARY - 
`The goal of this study is to predict the manner in which exercises were performed. The outcome designates the quality of exercise performed by alphabets A through E. To achieve this objective the training data set was used to build a prediction model which compares the results to the output variable classe. Some key milestones include
1. performing exploratory data analysis to filter out the needed variables,
2. partitioning the training data set into sub-train and sub-test pieces,
3. performing cross validation and estimating error/accuracy,
4. evaluating different models (rpart, lda, gbm, nb, rf) and choosing the best based on accuracy, and
5. generating responses to the final test data provided by Coursera




EXPLORATORY DATA ANALYSIS - 
The training data set was initially analyzed for existense of NA values and character values. Once determined that these were not needed the columns were removed. The original 160 columns were reduced to a much smaller set containing only 50 columns

```{r, exploratory data analysis, echo=FALSE }
        # Download training file in case it's not on current directory
            down_load_train_file <- "pml-training.csv"
            url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
            if( !file.exists(down_load_train_file) ) download.file(url, down_load_train_file, method="curl")
            if( !file.exists(down_load_train_file) ) download.file(url, down_load_train_file, method="internal")
            train_data <- read.csv( down_load_train_file, header=TRUE); # dim( train_data )
            
        # Download training file in case it's not on current directory
            down_load_test_file <- "pml-testing.csv"
            url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
            if( !file.exists(down_load_test_file) ) download.file(url, down_load_test_file, method="curl")
            if( !file.exists(down_load_test_file) ) download.file(url, down_load_test_file, method="internal")
            test_data <- read.csv( down_load_test_file, header=TRUE); # dim( test_data )
    
        # Check for NA values
            # table( colSums(is.na(train_data)) )
        # Truncate data set to smaller columns
            choola <- grep( 
        "^kur|^skew|^avg|^var|^std|^max|^min|^amp|timestamp|window|user_name|roll_forearm|pitch_forearm|yaw_forearm|X", 
            names(train_data), value=TRUE)
            myvars <- names(train_data) %in% choola
            test_selects <- test_data[!myvars]; # dim( test_selects )
            train_selects <- train_data[!myvars]; # dim( train_selects )

        # Do basic correlation analysis
            MM <- abs( cor(train_selects[,-50]))
            diag(MM) <- 0
            # which( MM > 0.8, arr.ind=T)

            cat("dimensions of original training data set =", dim(train_data) )
            cat( "dimensions of reduced training data set =", dim(train_selects))

````




PARTITION DATA - 
The training data set is partitioned into a training and test piece used for training the models and development of error/accuracy estimates

```{r, data partitioning, echo=FALSE}
        # partition data
            train_partitioned <- createDataPartition( y=train_selects$classe, p=0.75, list=FALSE)
            train_train <- train_selects[ train_partitioned,]
            train_test <- train_selects[ -train_partitioned,]

        cat( "dimensions of training piece of partitioned training set = ",  dim(train_train) )
        cat( "dimensions of test piece of partitioned training set = ",  dim(train_test) )
```



OUT OF SAMPLE ERROR/ACCURACY ESTIMATES - 
Two approaches are taken here. First, the partitioned training data is used to get an estimate of the error/accuracy. 

```{r, accuracy estimation, echo=FALSE, warning=FALSE, message=FALSE}
                set.seed(1234)
            # try "rpart" - 60 % on train data, 58 % on test data
                m1 <- train( classe ~., method="rpart", data=train_train )
                rpart_train <- confusionMatrix( train_train$classe, predict(m1, train_train))
                rpart_test <- confusionMatrix( train_test$classe, predict(m1, train_test))
            
            # try "lda" - 70 % on train data, 69 % on test data
                m2 <- train( classe ~., method="lda", data=train_train )
                lda_train <- confusionMatrix( train_train$classe, predict(m2, train_train))
                lda_test <- confusionMatrix( train_test$classe, predict(m2, train_test))
            
            # try "gbm" - 96 % on train data, 95 % on test data ( 22 min run time )
                m3 <- train( classe ~., method="gbm", data=train_train, verbose=FALSE )
                gbm_train <- confusionMatrix( train_train$classe, predict(m3, train_train))
                gbm_test <- confusionMatrix( train_test$classe, predict(m3, train_test))
                    
                cat("NOTE - Please interpret Accuracies as Accuracy = 1 - Error")
                cat("Model RPART Accuracy on train data=", round(rpart_train$overall[1],2),"& Accuracy on test data=",round(rpart_test$overall[1],2) )
            cat("Model LDA Accuracy on train data=", round(lda_train$overall[1],2),"& Accuracy on test data=",round(lda_test$overall[1],2) )
            cat("Model GBM Accuracy on train data=", round(gbm_train$overall[1],2),"& Accuracy on test data=", round(gbm_test$overall[1],2))

```



CROSS VALIDATION - 
Second, the trainControl() function was used to perform a repeated, k-fold cross validaton
```{r, cross validation, echo=FALSE, warning=FALSE, message=FALSE}

            # try "gbm" on full set - 96 % on train data, 95 % on test data ( 22 min run time, hence not running )
             #   m33 <- train( classe ~., method="gbm", data=train_selects )
             #   confusionMatrix( train_selects$classe, predict(m33, train_selects))
             #   answer_1 <- predict( m33, newdata = test_data)
            
            # try "gbm" on full set - using train control
             train_control <- trainControl( method="repeatedcv", number=10, repeats=3 )
             m34 <- train( classe ~., method="gbm", data=train_selects, trControl=train_control, verbose=FALSE)
             # confusionMatrix( train_selects$classe, predict(m34, train_selects))
             
    
        cat(" Model GBM with repeated k-fold cross validation on entire training data set gives Accuracies (1-Error)")
        cat(" values=", round( m34$results$Accuracy, 2 ) )
```



MODEL SELECTION STRATEGY - 
Based on accuracies generated on applying different models, the GBM model was chosen to evaluate the Project test case. 

```{r, model application to test case, echo=FALSE}
    
        answer <- predict( m34, newdata = test_data)

```



GENERATING OUTPUT FILE - Based on script provided by Project writeups an output file is generated

```{r, generate output file, echo=FALSE }
                pml_write_files = function(x){
                    n = length(x)
                    for(i in 1:n){
                        filename = paste0("problem_id_",i,".txt")
                        write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
                    }
                }
            
                # pml_write_files(answer)
```



CONCLUSIONS - The GBM model gives excellent results - 100 %. All the test cases provided in the Course Project Submission evaluated correctly.